What We Eval in the Shadows

### 2021-04-06 paper/main.tex

\subsection{Provenance.}
% Explain tha some generate expression objects and other language objects?
There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
    \item \c{substitute} synthesizes ASTs from expressions by replacing
    symbols with their bindings in the specified environment.
    \item \c{match.call} captures a call and its arguments as expressions.
    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
    \item \c{expression}  creates a vector of expression
    objects from text.
    \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
    expressions.
    \item \c{comp} compiles R code into R bytecode.
\end{itemize}

\begin{table}[h]
  \begin{tabular}{|c|c|c|}
      \hline
      & \#sites & \%sites \\
       \hline
         \c{substitute}, \c{quote}, ... & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
      \hline
      \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
      \hline
        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\   
      \hline
      \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
      \hline
      \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
      \hline
      \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
      \hline
  \end{tabular}
\caption{Provenances of \eval expressions. }
\end{table}

% \packageUnknownProvSitePercent : provenance not relavant (values, symbols etc)

\c{eval(parse(...))} can be used for dynamic code loading. This forms
the core of \c{source} and \c{sys.source} functions in R that are
commonly used for loading code in R files in interactive settings. We
investigated the number of cases in which the output of \c{parse} and
its variants is passed to \eval, directly or transitively by tainting
their output.  We observed that very few of the eval call (\packageNbParseFromFileSites in total)
consume the result of calling \c{parse} on a file. Most of the eval
calls consume the result of calling \c{parse} on a string. We also
identified one function in core R, \c{invokeRestartInteractively} that
prompts the user for input, parses it, and passes it to \eval.

@@

\todo{Illustrate things that we have discussed in Section 5 - examples}
\todo{Things to discuss: model.frame, match.call, substitute, parse, plots}

>>>>

\subsection{Provenance.} 
% Explain that some generate expression objects and other language objects?
There are various ways to obtain expressions that \eval will evaluate:
\begin{itemize}
    \item \c{substitute} synthesizes ASTs from expressions by replacing
    symbols with their bindings in the specified environment.
    \item \c{match.call} captures a call and its arguments as expressions.
    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression. % Aldd \c{call} ?
    \item \c{expression}  creates a vector of expression
    objects from text.
    \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
    expressions.
    \item \c{comp} compiles R code into R bytecode.
\end{itemize}
We show the distribution of the different provenances in Table~\ref{tab:provenance}. Note that one site could be passed arguments from different provenances depending on the invokation. For some arguments, we could not determine their provenance or it was a value or a symbol, for which the provenance is less significant. % Not clear... Or maybe add symbols in that table?
% One can create a symbol with as.name, as.symbol etc. That would be the provenance (but a symbol can also come from quote for instance...)

\begin{table}[h]
    \begin{tabular}{|c|c|c|}
        \hline
        & \#sites & \%sites \\
        \hline
        \c{substitute}, \c{quote}, ... & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
        \hline
        \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
        \hline
        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\   
        \hline
        \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
        \hline
        \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
        \hline
        \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
        \hline \hline 
        Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\
        \hline
    \end{tabular}
    \caption{Provenances of \eval expressions. } \label{tab:provenance}
\end{table}

We taint the result of a call to \c{parse}, \c{str2lang} and \c{str2expression}, by attaching an R attribute to it. R attributes are lost under copy or modification of their attaching object, so we underestimate the number of sites in that case. However, we observed that most of the time, the result of \c{parse} is not touched and directly passed to \eval.

We taint the result of \c{match.call} by remembering the addresses of all the elements of the call list. As R GC is non-moving, it makes it possible to track them individually, as we observed that calls are modified and individual elements are sometimes evaluated. This is also an underestimation of the number of sites though, as we would lose an individual element of the call list that would be modified.

% explain also how we get the numbers of the other categories?

\c{eval(parse(...))} (and \c{str2expression}, and \c{str2lang}) can be used for dynamic code loading. This forms
the core of \c{source} and \c{sys.source} functions in R that are
commonly used for loading code in R files in interactive settings. We
investigated the number of cases in which the output of \c{parse} and
its variants is passed to \eval, directly or transitively by tainting
their output.  We observed that very few of the eval call (\packageNbParseFromFileSites in total)
consume the result of calling \c{parse} on a file. Most of the eval
calls consume the result of calling \c{parse} on a string. We also
identified one function in core R, \c{invokeRestartInteractively} that
prompts the user for input, parses it, and passes it to \eval.

In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed  an expressions that result from a call to \c{parse}.

@@

\todo{Illustrate things that we have discussed in Section 5 - examples}
\todo{Things to discuss: model.frame, match.call, subset as example of NSE, substitute, parse, plots}
